Adaptive Video Streaming with PID-Based Quality Control and Optimized Bitstream Extraction

### 2015-02-06 tmm.tex
changed the abstract, before the introduction

% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
\author{Shengbin~Meng,~Jun~Sun,~Yizhou~Duan~and~Zongming~Guo % <-this % stops a space
\thanks{S. Meng is with the Institute of Computer Science and Technology, Peking University, Beijing%
, China, 100080. E-mail: shengbin@pku.edu.cn.}% <-this % stops a space
\thanks{J. Sun, Y. Duan, Z. Guo are with the Institute of Computer Science and Technology, Peking University.}% <-this % stops a space
}%\thanks{Manuscript received September 15, 2014}}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
In order to meet the challenge brought by bandwidth fluctuation of the user's network, modern video streaming systems tend to be adaptive, i.e., able to adjust the video quality according to available bandwidth. This paper implements such an adaptive video streaming system which utilizes the Scalable Video Coding (SVC) technology, and gives novel solutions for two key problems within the system. First, a quality control algorithm is proposed for the video streaming system to adjust its video quality along with the real bandwidth change. The algorithm is based on the Proportional-Integral-Derivative (PID) control method and designed to deliver best video quality with least quality fluctuation. Second, an efficient solution is present for the bitstream extraction problem, which aims to select an optimal sub-stream from the whole SVC video bitstream under given bitrate constraint. The solution features a linear error model to estimate the distortion caused by discarding any combination of data packets, and a greedy algorithm scheme to assign each data packet a priority value for Rate-Distortion optimized extraction. Experimental results show that the PID-based quality control algorithm can drastically reduce the quality fluctuation while still keeping a high quality level, and the bitstream extraction method can achieve a significant PSNR gain compared to existing reference extractors, without computational complexity increment. The system is currently running at the online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the client to keep a non-paused smooth playing. Therefore, to achieve better watching experience for users with various network environment, the video's bitrate or quality should be able to adjust according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, which allows efficient scalability of video bitrate and quality with only one single bitstream. The generally-known term SVC usually refers to the scalable extension of the H.264/AVC video coding standard \cite{SVCOverview}. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into multiple layers: one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most existing video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity simulcast provides is strictly limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. An SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), and thus can provide seamless scalability and adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, SVC is considered to be a better choice for adaptive video streaming systems. While our work in this paper can be partly used in a simulcast system, we will mainly focus on building an adaptive streaming system using SVC.

To guarantee good performance of SVC streaming system, there are two important issues that need to be addressed well. First is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (as well as the bitrate) to follow the change of bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constrain of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

In this paper, we propose novel solutions for these two issues and construct an efficient adaptive video streaming system. Apple Darwin Streaming Server (DSS) \cite{DSS}, which has been widely used to collect, cache, schedule and transmit multimedia contents in practice, is chosen as the basis of the system. We modified the original DSS project and added SVC support to it, for the purpose of both research and application. A new quality control algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves higher average quality level with less quality fluctuation. For the bitstream extraction problem, we propose a linear error model to directly estimate pixel value errors before MSE or PSNR is calculated. This model enables us to accurately estimate the distortion caused by discarding any combination of packets. Based on that, we design a greedy algorithm to assign priority for each quality enhancement packet according to its rate-distortion (R-D) impact. The priority value of each packet can be stored in the bitstream and used for optimized extraction. The proposed algorithms perform well in experiments and the quality control algorithm is also implemented in the practical streaming system we construct and proved quite effective.

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related works on them. Then in section \ref{sec:quality-control}, the proposed quality control algorithm is formulated and explained with term definitions and parameters selection. After that, we propose the solution for the bitstream extraction problem in section \ref{sec:bit-extraction}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimized Bitstream Extraction Based On Linear Error Model}
\label{sec:bit-extraction}

@@

\subsection{Distortion Estimation And Model Verification}

@@

In this paper, we propose an adaptive video streaming system with PID-based quality control algorithm and optimized bitstream extraction. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The bitstream extraction method we propose includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The packets' priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5 compared to the JSVM QL extractor under the same bitrate constrain, also with the same computational complexity.

>>>>

% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
\author{Shengbin~Meng,~Jun~Sun,~Yizhou~Duan~and~Zongming~Guo,~\em{Member,~IEEE} % <-this % stops a space
\thanks{Manuscript received Febrary 15, 2015.}
\thanks{The authors are with the Institute of Computer Science and Technology, Peking University, Beijing%
, 100080, China (e-mail: shengbin@pku.edu.cn; sunjun@pku.edu.cn; duanyizhou@pku.edu.cn; guozongming@pku.edu.cn).}}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This paper proposes an adaptive video streaming system that adjusts the video quality according to available bandwidth and gives the user best quality of service under various network conditions. Two key problems within the system are analyzed and solved with novel algorithms. First, we utilize the Proportional-Integral-Derivative (PID) method in video quality control to determine a suitable bitrate for transmission, so that the system can deliver highest video quality with least quality fluctuation. Second, to select an optimal sub-stream for transmission under given bitrate constraint, we use a linear error model to estimate the distortion caused by discarding any combination of data packets, and a greedy algorithm scheme to assign each data packet a priority value for Rate-Distortion optimized extraction. Experimental results show that the PID-based quality control algorithm brings 8.6\% higher video quality as well as 24.8\% lower quality fluctuation than the baseline algorithm, and the bitstream extraction method achieves a PSNR gain of up to 0.5dB compared to existing reference extractors, with no computational complexity increment. The system is currently running at the online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

One of the biggest challenge for video streaming is the variety of users' network environment. A user may use different types of network (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another (e.g., from Wi-Fi to cellular network) on going. Even in the same network, the user's bandwidth may change from time to time, influenced by the overall traffic. During such an environment, bandwidth fluctuation is inevitable. If the video is streamed at a constant bitrate, it would be difficult or impossible for the user to get best watching experience. To cope with the challenge, modern video streaming systems should be able to automatically adjust the video's bitrate or quality according to the bandwidth change. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams with different video quality are stored at the server, and the client selects one of them according to currently available bandwidth. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, which allows scalability of video bitrate and quality with only one single bitstream. The generally-known term SVC usually refers to the scalable extension of the H.264/AVC video coding standard \cite{SVCOverview}. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into multiple layers: one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate data rate and video quality as required.

The advantage of simulcast systems is that they need little modification of existing infrastructure and therefore are easy to deploy. Most existing video websites, YouTube \cite{YouTube} for instance, use this option to enable quality adaptation. However, the adaptive range and granularity simulcast provides is strictly limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams significantly increase time and space cost. SVC streaming systems, on the contrary, do not have above shortcomings. Since an SVC video bitstream can be extracted at the granularity of one NALU (a very small partition of data), it is able to provide seamless scalability and quality adaptation. Besides, performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, despite some computation and bitrate overhead comparing with single layer stream. Therefore, SVC is considered to be a better choice for adaptive video streaming systems. While the work of this paper can be partly used in a simulcast system, it will mainly focus on the adaptive streaming system using SVC.

In a video streaming system using SVC, there are two key problems that need to be addressed well. The first is quality control, and the second is bitstream extraction. Quality control means controlling the video quality (or the video bitrate, from the transmission point of view) to follow the change of the bandwidth. Once the bandwidth has changed, the quality control algorithm should determine a bitrate suitable for the current bandwidth. Given this bitrate, how to extract data from the whole SVC stream to form an optimal sub-stream becomes the bitstream extraction problem. Optimized bitstream extraction aims to select a subset of NALUs (or packets) which represents video with best quality, or minimal distortion, under the constrain of the given bitrate. It should be noted that, bitstream extraction is not only useful in adaptive video streaming, but also in other SVC application scenarios where part of data needs to be obtained or kept, e.g., the archiving of surveillance video.

In this paper, an efficient adaptive video streaming system based on SVC is proposed. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the system. The original DSS project is modified by us and SVC support is added to it, for the purpose of both research and application. Then novel algorithms are designed and implemented to better solve the two key problems mentioned above. The contributions of our work are threefold:
\begin{itemize}
\item A quality control algorithm based on the Proportional-Integral-Derivative (PID) control concept is proposed to replace the native one in DSS, and achieves higher average quality level with less quality fluctuation.
\item A linear error model is proposed to accurately estimate the distortion caused by discarding any combination of packets from a SVC bitstream.
\item A greedy-like algorithm is designed to assign priority for each packet, thus enables optimized bitstream extraction.
\end{itemize}

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related works on them. Then in section \ref{sec:quality-control}, the proposed quality control algorithm is formulated and explained with term definitions and parameters selection. After that, we present the solution for the bitstream extraction problem in section \ref{sec:bit-extraction}, including the linear error model and the greed-like algorithm. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimized Bitstream Extraction Based on Linear Error Model}
\label{sec:bit-extraction}

@@

\subsection{Distortion Estimation and Model Verification}

@@

In this paper, we propose an adaptive video streaming system with PID-based quality control algorithm and optimized bitstream extraction. The PID-based quality control algorithm has a flexible mechanism to determine the suitable quality level so that it gives an accurate and swift response to the bandwidth change. Comparing with the original algorithm in DSS, this algorithm has reduced quality fluctuation significantly to improve audience satisfaction. The bitstream extraction method we propose includes a linear error model to estimate the distortion caused by discarding any combination of packets and a greedy algorithm to assign priority to each data packet. The packets' priority values can then be stored in the bitstream and used for R-D optimized extraction. Experimental results show that the proposed method can achieve PSNR gain of up to 0.5dB compared to the JSVM QL extractor under the same bitrate constrain, also with the same computational complexity.
